---
title: "NCMP Ethnic-specific BMI Adjustment Report"
date: last-modified
date-format: "MMM YYYY"
format: 
  html:
    grid:
      body-width: 900px
    embed-resources: true
toc: true
editor: source
execute: 
  echo: false
  warning: false
---

```{r setup}
library(tidyverse)
library(readxl)
library(plotly)
library(PHEindicatormethods)
library(bcctheme)
library(ggiraph)

BMI_Category_Levels = c("Underweight", "Healthy Weight", "Overweight", "Obese")
bcc_colours <- unname(bcc_cols())
BMI_Category_colours <- c("Underweight" = bcc_colours[6],
                          "Healthy Weight" = bcc_colours[4],
                          "Overweight" = bcc_colours[5],
                          "Obese" = bcc_colours[1]) 

# function to round an integer to nearest 5
# for data suppression purposes: https://digital.nhs.uk/data-and-information/find-data-and-publications/statement-of-administrative-sources/methodological-changes/changes-to-the-national-child-measurement-programme-2019-20-publication-content-and-disclosure-control-methodology#simple-calculations
round_5 <- function(x) {
  return(round(x / 5) * 5)
}

# function to summarise a patient level dataset by the columns given
# .by_count contains columns to summarise by
# .by_denom contains columns to create denominator
# returns a dataframe containing count, denominator and percentage with 95% CIs
# optionally include overweight/obese category (for BMI_Category data only)
# optionally recalculate percentage using count and denom rounded to nearest 5
ncmp_summarise <- function(.data, 
                           .by_count, 
                           .by_denom, 
                           .include_oo,
                           .round_5){
  if(.include_oo == TRUE){
    temp <- .data |> 
      dplyr::summarise(count = n(), .by = {{ .by_count }}) |> 
      dplyr::mutate(denominator = sum(count), .by = {{ .by_denom }})
    
    temp_oo <- .data |> 
      dplyr::mutate(BMI_Category = dplyr::case_when(BMI_Category == "Obese" ~ "Overweight/Obese",
                                      BMI_Category == "Overweight" ~ "Overweight/Obese",
                                      .default = BMI_Category)) |> 
      dplyr::summarise(count = n(), .by = {{ .by_count }}) |> 
      dplyr::mutate(denominator = sum(count), .by = {{ .by_denom }}) |> 
      dplyr::filter(BMI_Category == "Overweight/Obese")
    
    temp <- rbind(temp, temp_oo)
    
    if(.round_5 == TRUE){
      temp <- PHEindicatormethods::phe_proportion(temp,
                                                  x = count,
                                                  n = denominator,
                                                  multiplier = 100) |> 
        dplyr::mutate(value = (round_5(count)/round_5(denominator))*100) |> 
        dplyr::rename(pct = value,
               lower_ci = lowercl,
               upper_ci = uppercl)
    }
    
    if(.round_5 == FALSE){
      temp <- PHEindicatormethods::phe_proportion(temp,
                                                  x = count,
                                                  n = denominator,
                                                  multiplier = 100) |> 
        dplyr::rename(pct = value,
               lower_ci = lowercl,
               upper_ci = uppercl)
    }
  }
  
  if(.include_oo == FALSE){
    temp <- .data |> 
      dplyr::summarise(count = n(), .by = {{ .by_count }}) |> 
      dplyr::mutate(denominator = sum(count), .by = {{ .by_denom }})
    
    if(.round_5 == TRUE){
      temp <- PHEindicatormethods::phe_proportion(temp,
                                                  x = count,
                                                  n = denominator,
                                                  multiplier = 100) |> 
        dplyr::mutate(value = (round_5(count)/round_5(denominator))*100) |> 
        dplyr::rename(pct = value,
               lower_ci = lowercl,
               upper_ci = uppercl)
    }
    
    if(.round_5 == FALSE){
      temp <- PHEindicatormethods::phe_proportion(temp,
                                                  x = count,
                                                  n = denominator,
                                                  multiplier = 100) |> 
        dplyr::rename(pct = value,
               lower_ci = lowercl,
               upper_ci = uppercl)
    }
  }
  temp
}

# function to generate an interactive sankey diagram showing movement between BMI classes
# .Ethnicity_Hudda is ethnicity to plot
#. School_Year is school year to plot
bmi_adj_sankey <- function(.data,
                           .Ethnicity_Hudda,
                           .School_Year){
  
  sankey_links <- .data |>
    dplyr::filter(Ethnicity_Hudda == .Ethnicity_Hudda,
           School_Year == .School_Year) |> 
    dplyr::select(BMI_Category, BMI_Category_Adjusted, count, pct, lower_ci, upper_ci) |>
    dplyr::rename(source = BMI_Category,
           target = BMI_Category_Adjusted,
           value  = count) |>
    dplyr::mutate(
      linkgroup = source,
      linklabel = dplyr::case_when(
        source == "Underweight"     & target == "Underweight"     ~ "Remained underweight",
        source == "Healthy Weight"  & target == "Healthy Weight"  ~ "Remained healthy weight",
        source == "Overweight"      & target == "Overweight"      ~ "Remained overweight",
        source == "Obese"           & target == "Obese"           ~ "Remained obese",
        .default = paste0("Moved to ", str_to_lower(target))),
      source = paste0(source, "_unadj"),
      target = paste0(target, "_adj"),
      colour = dplyr::case_when(
        linkgroup == "Underweight"    ~ paste0(BMI_Category_colours["Underweight"], "99"),
        linkgroup == "Healthy Weight" ~ paste0(BMI_Category_colours["Healthy Weight"], "99"),
        linkgroup == "Overweight"     ~ paste0(BMI_Category_colours["Overweight"], "99"),
        linkgroup == "Obese"          ~ paste0(BMI_Category_colours["Obese"], "99"))) |>
    dplyr::filter(value > 7)
  
  # Explicit node order (left then right)
  ordered_nodes <- c("Underweight_unadj", "Healthy Weight_unadj", "Overweight_unadj", "Obese_unadj",
                     "Underweight_adj", "Healthy Weight_adj", "Overweight_adj", "Obese_adj")
  
  # Build nodes with consistent labels and colours
  sankey_nodes <- tibble::tibble(name = ordered_nodes,
                         # category without suffix, for color mapping
                         nodegroup = sub("_(unadj|adj)$", "", ordered_nodes),
                         colour = dplyr::case_when(
                           nodegroup == "Underweight"    ~ BMI_Category_colours["Underweight"],
                           nodegroup == "Healthy Weight" ~ BMI_Category_colours["Healthy Weight"],
                           nodegroup == "Overweight"     ~ BMI_Category_colours["Overweight"],
                           nodegroup == "Obese"          ~ BMI_Category_colours["Obese"]))
  
  # Labels for nodes
  node_labels <- sub("_(unadj|adj)$", "", sankey_nodes$name)
  
  # Map links to node indices based on explicit node order
  sankey_links$IDsource <- match(sankey_links$source, sankey_nodes$name) - 1
  sankey_links$IDtarget <- match(sankey_links$target, sankey_nodes$name) - 1
  
  # Plot
  plotly::plot_ly(type = "sankey",
          arrangement = "snap",
          orientation = "h",
          node = list(label = node_labels,          # length 8; no recycling
                      color = sankey_nodes$colour,
                      pad = 15,
                      thickness = 20,
                      line = list(color = "black", width = 0.5),
                      hoverinfo = "none"),
          link = list(source = sankey_links$IDsource,
                      target = sankey_links$IDtarget,
                      value  = sankey_links$value,
                      label  = sankey_links$linklabel,
                      color  = sankey_links$colour,
                      customdata = sprintf("%.1f%% (%.1f%% - %.1f%%)", sankey_links$pct, sankey_links$lower_ci, sankey_links$upper_ci),
                      hovertemplate = paste0("%{label}<br>%{customdata}<extra></extra>"))) |> 
    add_annotations(x = 0, y = 1.01, xref = "paper", yref = "paper",
                    text = "Unadjusted BMI score", showarrow = FALSE,
                    xanchor = "left", yanchor = "bottom",
                    font = list(size = 16, color = "black")) |> 
    add_annotations(x = 1, y = 1.01, xref = "paper", yref = "paper",
                    text = "Adjusted BMI score", showarrow = FALSE,
                    xanchor = "right", yanchor = "bottom",
                    font = list(size = 16, color = "black")) |> 
    layout(margin = list(t = 40)) |> 
    config(modeBarButtonsToRemove = c('zoom', 'pan', 'select', 'zoomIn', 'zoomOut', 'lasso2d', 'autoScale', 'resetScale'))
}

bmi_adj_sankey_clinical <- function(.data,
                           .Ethnicity_Hudda,
                           .School_Year){
  
  sankey_links <- .data |>
    dplyr::filter(Ethnicity_Hudda == .Ethnicity_Hudda,
           School_Year == .School_Year) |> 
    dplyr::select(BMI_Category_Clinical, BMI_Category_Clinical_Adjusted, count, pct, lower_ci, upper_ci) |>
    dplyr::rename(source = BMI_Category_Clinical,
           target = BMI_Category_Clinical_Adjusted,
           value  = count) |>
    dplyr::mutate(
      linkgroup = source,
      linklabel = dplyr::case_when(
        source == "Underweight"     & target == "Underweight"     ~ "Remained underweight",
        source == "Healthy Weight"  & target == "Healthy Weight"  ~ "Remained healthy weight",
        source == "Overweight"      & target == "Overweight"      ~ "Remained overweight",
        source == "Obese"           & target == "Obese"           ~ "Remained obese",
        .default = paste0("Moved to ", str_to_lower(target))),
      source = paste0(source, "_unadj"),
      target = paste0(target, "_adj"),
      colour = dplyr::case_when(
        linkgroup == "Underweight"    ~ paste0(BMI_Category_colours["Underweight"], "99"),
        linkgroup == "Healthy Weight" ~ paste0(BMI_Category_colours["Healthy Weight"], "99"),
        linkgroup == "Overweight"     ~ paste0(BMI_Category_colours["Overweight"], "99"),
        linkgroup == "Obese"          ~ paste0(BMI_Category_colours["Obese"], "99"))) |>
    dplyr::filter(value > 7)
  
  # Explicit node order (left then right)
  ordered_nodes <- c("Underweight_unadj", "Healthy Weight_unadj", "Overweight_unadj", "Obese_unadj",
                     "Underweight_adj", "Healthy Weight_adj", "Overweight_adj", "Obese_adj")
  
  # Build nodes with consistent labels and colours
  sankey_nodes <- tibble::tibble(name = ordered_nodes,
                         # category without suffix, for color mapping
                         nodegroup = sub("_(unadj|adj)$", "", ordered_nodes),
                         colour = dplyr::case_when(
                           nodegroup == "Underweight"    ~ BMI_Category_colours["Underweight"],
                           nodegroup == "Healthy Weight" ~ BMI_Category_colours["Healthy Weight"],
                           nodegroup == "Overweight"     ~ BMI_Category_colours["Overweight"],
                           nodegroup == "Obese"          ~ BMI_Category_colours["Obese"]))
  
  # Labels for nodes
  node_labels <- sub("_(unadj|adj)$", "", sankey_nodes$name)
  
  # Map links to node indices based on explicit node order
  sankey_links$IDsource <- match(sankey_links$source, sankey_nodes$name) - 1
  sankey_links$IDtarget <- match(sankey_links$target, sankey_nodes$name) - 1
  
  # Plot
  plotly::plot_ly(type = "sankey",
          arrangement = "snap",
          orientation = "h",
          node = list(label = node_labels,          # length 8; no recycling
                      color = sankey_nodes$colour,
                      pad = 15,
                      thickness = 20,
                      line = list(color = "black", width = 0.5),
                      hoverinfo = "none"),
          link = list(source = sankey_links$IDsource,
                      target = sankey_links$IDtarget,
                      value  = sankey_links$value,
                      label  = sankey_links$linklabel,
                      color  = sankey_links$colour,
                      customdata = sprintf("%.1f%% (%.1f%% - %.1f%%)", sankey_links$pct, sankey_links$lower_ci, sankey_links$upper_ci),
                      hovertemplate = paste0("%{label}<br>%{customdata}<extra></extra>"))) |> 
    add_annotations(x = 0, y = 1.01, xref = "paper", yref = "paper",
                    text = "Unadjusted BMI score", showarrow = FALSE,
                    xanchor = "left", yanchor = "bottom",
                    font = list(size = 16, color = "black")) |> 
    add_annotations(x = 1, y = 1.01, xref = "paper", yref = "paper",
                    text = "Adjusted BMI score", showarrow = FALSE,
                    xanchor = "right", yanchor = "bottom",
                    font = list(size = 16, color = "black")) |> 
    layout(margin = list(t = 40)) |> 
    config(modeBarButtonsToRemove = c('zoom', 'pan', 'select', 'zoomIn', 'zoomOut', 'lasso2d', 'autoScale', 'resetScale'))
}
```

```{r data_import}
ncmp_data <- read_excel("../data/ncmp_data_adjusted_bmi.xlsx") |> 
  rename(BMI_z_Adjusted = SDS_BMI) |> 
  mutate(BMI_p_Adjusted = pnorm(BMI_z_Adjusted),
         BMI_Category_Adjusted = case_when(BMI_p_Adjusted >= 0.95 ~ "Obese",
                                           BMI_p_Adjusted >= 0.85 ~ "Overweight",
                                           BMI_p_Adjusted > 0.02 ~ "Healthy Weight",
                                           BMI_p_Adjusted <= 0.02 ~ "Underweight"),
         BMI_Category_Clinical_Adjusted = case_when(BMI_z_Adjusted >= 6/3 ~ "Obese",
                                                    BMI_z_Adjusted >= 4/3 ~ "Overweight",
                                                    BMI_z_Adjusted > -6/3 ~ "Healthy Weight",
                                                    BMI_z_Adjusted <= -6/3 ~ "Underweight"))

# Group into three year combined periods ----------------------------------

# put years into a vector
years <- unique(ncmp_data$Year)

# remove 2020/2021
years <- years[!years %in% c("2020/2021")]

# create df with three years of data (first three years i.e. oldest)
ncmp_data_period_1 <- ncmp_data |> 
  filter(Year == years[1]|Year == years[2]|Year == years[3]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[1]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[3])))

# same again but with next three year period
ncmp_data_period_2 <- ncmp_data |> 
  filter(Year == years[2]|Year == years[3]|Year == years[4]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[2]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[4])))

# and so on
ncmp_data_period_3 <- ncmp_data |> 
  filter(Year == years[3]|Year == years[4]|Year == years[5]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[3]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[5])))

ncmp_data_period_4 <- ncmp_data |> 
  filter(Year == years[4]|Year == years[5]|Year == years[6]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[4]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[6])))

ncmp_data_period_5 <- ncmp_data |> 
  filter(Year == years[5]|Year == years[6]|Year == years[7]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[5]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[7])))

ncmp_data_period_6 <- ncmp_data |> 
  filter(Year == years[6]|Year == years[7]|Year == years[8]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[6]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[8])))

ncmp_data_period_7 <- ncmp_data |> 
  filter(Year == years[7]|Year == years[8]|Year == years[9]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[7]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[9])))

# most recent period
ncmp_data_period_8 <- ncmp_data |> 
  filter(Year == years[8]|Year == years[9]|Year == years[10]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[8]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[10])))

# bind all into one df
ncmp_data_3_years_combined <- bind_rows(ncmp_data_period_1,
                                        ncmp_data_period_2,
                                        ncmp_data_period_3,
                                        ncmp_data_period_4,
                                        ncmp_data_period_5,
                                        ncmp_data_period_6,
                                        ncmp_data_period_7,
                                        ncmp_data_period_8)

periods <- unique(ncmp_data_3_years_combined$Year)

# get a list of periods in format 20XX-XX
# where most recent period is periods[8]
periods_short <- sub("/([A-Za-z0-9]+( [A-Za-z0-9]+)+)/", "-", unique(ncmp_data_3_years_combined$Year))
```

```{r create_dfs}
# Find prevalences before and after adjustment (population) ----------------------------

# summarise unadjusted BMI categories by school year

bmi_category_unadj_by_school_year <- ncmp_summarise(ncmp_data_period_8,
                                                  .by_count = c(School_Year, BMI_Category),
                                                  .by_denom = c(School_Year),
                                                  .include_oo = T,
                                                  .round_5 = T) |> 
  mutate(bmi_adj = "Unadjusted BMI")

# summarise adjusted BMI categories by ethnicity

bmi_category_adj_by_school_year <- ncmp_data_period_8 |> 
  select(-BMI_Category) |> 
  rename(BMI_Category = BMI_Category_Adjusted) |> 
  ncmp_summarise(.by_count = c(School_Year, BMI_Category),
                 .by_denom = c(School_Year),
                 .include_oo = T,
                 .round_5 = T) |> 
  mutate(bmi_adj = "Adjusted BMI")

# combine into one table

bmi_category_by_school_year = rbind(bmi_category_adj_by_school_year,
                                  bmi_category_unadj_by_school_year)

# pull out overall overweight/obese % before and after adjustment
oo_r_unadj <- bmi_category_by_school_year |> 
  filter(bmi_adj == "Unadjusted BMI",
         School_Year == "Reception",
         BMI_Category == "Overweight/Obese") |> 
  pull(pct) |> 
  round(1)

oo_y6_unadj <- bmi_category_by_school_year |> 
  filter(bmi_adj == "Unadjusted BMI",
         School_Year == "Year 6",
         BMI_Category == "Overweight/Obese") |> 
  pull(pct) |> 
  round(1)

oo_r_adj <- bmi_category_by_school_year |> 
  filter(bmi_adj == "Adjusted BMI",
         School_Year == "Reception",
         BMI_Category == "Overweight/Obese") |> 
  pull(pct) |> 
  round(1)

oo_y6_adj <- bmi_category_by_school_year |> 
  filter(bmi_adj == "Adjusted BMI",
         School_Year == "Year 6",
         BMI_Category == "Overweight/Obese") |> 
  pull(pct) |> 
  round(1)

# summarise unadjusted BMI categories by ethnicity

bmi_category_unadj_by_ethnicity <- ncmp_summarise(ncmp_data_period_8,
                                                  .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category),
                                                  .by_denom = c(School_Year, Ethnicity_Hudda),
                                                  .include_oo = T,
                                                  .round_5 = T) |> 
  mutate(bmi_adj = "Unadjusted BMI")

# summarise adjusted BMI categories by ethnicity

bmi_category_adj_by_ethnicity <- ncmp_data_period_8 |> 
  select(-BMI_Category) |> 
  rename(BMI_Category = BMI_Category_Adjusted) |> 
  ncmp_summarise(.by_count = c(School_Year, Ethnicity_Hudda, BMI_Category),
                 .by_denom = c(School_Year, Ethnicity_Hudda),
                 .include_oo = T,
                 .round_5 = T) |> 
  mutate(bmi_adj = "Adjusted BMI")

# combine into one table

bmi_category_by_ethnicity = rbind(bmi_category_adj_by_ethnicity,
                                  bmi_category_unadj_by_ethnicity)

# pull out overweight/obese % before and after adjustment
oo_r_unadj_b <- bmi_category_by_ethnicity |> 
  filter(bmi_adj == "Unadjusted BMI",
         School_Year == "Reception",
         BMI_Category == "Overweight/Obese",
         Ethnicity_Hudda == "Black") |> 
  pull(pct) |> 
  round(1)

oo_y6_unadj_b <- bmi_category_by_ethnicity |> 
  filter(bmi_adj == "Unadjusted BMI",
         School_Year == "Year 6",
         BMI_Category == "Overweight/Obese",
         Ethnicity_Hudda == "Black") |> 
  pull(pct) |> 
  round(1)

oo_r_adj_b <- bmi_category_by_ethnicity |> 
  filter(bmi_adj == "Adjusted BMI",
         School_Year == "Reception",
         BMI_Category == "Overweight/Obese",
         Ethnicity_Hudda == "Black") |> 
  pull(pct) |> 
  round(1)

oo_y6_adj_b <- bmi_category_by_ethnicity |> 
  filter(bmi_adj == "Adjusted BMI",
         School_Year == "Year 6",
         BMI_Category == "Overweight/Obese",
         Ethnicity_Hudda == "Black") |> 
  pull(pct) |> 
  round(1)

oo_r_unadj_sa <- bmi_category_by_ethnicity |> 
  filter(bmi_adj == "Unadjusted BMI",
         School_Year == "Reception",
         BMI_Category == "Overweight/Obese",
         Ethnicity_Hudda == "South Asian") |> 
  pull(pct) |> 
  round(1)

oo_y6_unadj_sa <- bmi_category_by_ethnicity |> 
  filter(bmi_adj == "Unadjusted BMI",
         School_Year == "Year 6",
         BMI_Category == "Overweight/Obese",
         Ethnicity_Hudda == "South Asian") |> 
  pull(pct) |> 
  round(1)

oo_r_adj_sa <- bmi_category_by_ethnicity |> 
  filter(bmi_adj == "Adjusted BMI",
         School_Year == "Reception",
         BMI_Category == "Overweight/Obese",
         Ethnicity_Hudda == "South Asian") |> 
  pull(pct) |> 
  round(1)

oo_y6_adj_sa <- bmi_category_by_ethnicity |> 
  filter(bmi_adj == "Adjusted BMI",
         School_Year == "Year 6",
         BMI_Category == "Overweight/Obese",
         Ethnicity_Hudda == "South Asian") |> 
  pull(pct) |> 
  round(1)

# Find prevalences before and after adjustment (clinical) -----------------

# repeat for clinical categories

# summarise unadjusted BMI categories by ethnicity

bmi_category_clinical_unadj_by_ethnicity <- ncmp_summarise(ncmp_data_period_8,
                                                           .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical),
                                                           .by_denom = c(School_Year, Ethnicity_Hudda),
                                                           .include_oo = F,
                                                           .round_5 = T) |> 
  mutate(bmi_adj = "Unadjusted BMI")

# summarise adjusted BMI categories by ethnicity

bmi_category_clinical_adj_by_ethnicity <- ncmp_data_period_8 |> 
  ncmp_summarise(.by_count = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical),
                 .by_denom = c(School_Year, Ethnicity_Hudda),
                 .include_oo = F,
                 .round_5 = T) |> 
  mutate(bmi_adj = "Adjusted BMI")

# combine into one table

bmi_category_clinical_by_ethnicity = rbind(bmi_category_clinical_adj_by_ethnicity,
                                           bmi_category_clinical_unadj_by_ethnicity)

# Find % children changed category (population) ---------------------------

# calculate proportion of children within ethnicity/school year who fit into each BMI category/adjusted category combo
bmi_category_change_by_ethnicity <- ncmp_summarise(ncmp_data_period_8,
               .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category, BMI_Category_Adjusted),
               .by_denom = c(School_Year, Ethnicity_Hudda),
               .include_oo = F,
               .round_5 = T) |> 
  mutate(category_change = case_when(BMI_Category == BMI_Category_Adjusted ~ F,
                                     .default = T))

# add cols with number of "steps" and direction of change
bmi_category_change_by_ethnicity <- bmi_category_change_by_ethnicity |> 
  mutate(BMI_Category = factor(BMI_Category, levels = BMI_Category_Levels, ordered = T),
         BMI_Category_Adjusted = factor(BMI_Category_Adjusted, levels = BMI_Category_Levels, ordered = T),
         change_steps = as.numeric(unclass(BMI_Category_Adjusted) - unclass(BMI_Category)),
         change_direction = case_when(change_steps < 0 ~ "Lighter",
                                      change_steps > 0 ~ "Heavier",
                                      change_steps == 0 ~ "No change"))

# summarise to get % of all children in each ethnicity/school year who changed BMI category
bmi_category_change_by_ethnicity_summary <- bmi_category_change_by_ethnicity |> 
  filter(category_change == T) |> 
  group_by(School_Year, Ethnicity_Hudda) |> 
  summarise(count = sum(count),
            denominator = mean(denominator)) |> 
  ungroup() |> 
  PHEindicatormethods::phe_proportion(x = count,
                                      n = denominator,
                                      multiplier = 100)

# calculate proportion of children within ethncicity/school year/bmi category who fit into each adjusted category
adjusted_bmi_by_ethnicity_bmi_category <- ncmp_summarise(ncmp_data_period_8,
                                                         .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category, BMI_Category_Adjusted),
                                                         .by_denom = c(School_Year, Ethnicity_Hudda, BMI_Category),
                                                         .include_oo = F,
                                                         .round_5 = T) |> 
  mutate(category_change = case_when(BMI_Category == BMI_Category_Adjusted ~ F,
                                     .default = T))

# Find % children changed category (clinical) -----------------------------


# repeat for clinical category
# calculate proportion of children within ethnicity/school year who fit into each BMI category/adjusted category combo
bmi_category_clinical_change_by_ethnicity <- ncmp_summarise(ncmp_data_period_8,
                                                   .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical, BMI_Category_Clinical_Adjusted),
                                                   .by_denom = c(School_Year, Ethnicity_Hudda),
                                                   .include_oo = F,
                                                   .round_5 = T) |> 
  mutate(category_change = case_when(BMI_Category_Clinical == BMI_Category_Clinical_Adjusted ~ F,
                                     .default = T))

# add cols with number of "steps" and direction of change
bmi_category_clinical_change_by_ethnicity <- bmi_category_clinical_change_by_ethnicity |> 
  mutate(BMI_Category_Clinical = factor(BMI_Category_Clinical, levels = BMI_Category_Levels, ordered = T),
         BMI_Category_Clinical_Adjusted = factor(BMI_Category_Clinical_Adjusted, levels = BMI_Category_Levels, ordered = T),
         change_steps = as.numeric(unclass(BMI_Category_Clinical_Adjusted) - unclass(BMI_Category_Clinical)),
         change_direction = case_when(change_steps < 0 ~ "Lighter",
                                      change_steps > 0 ~ "Heavier",
                                      change_steps == 0 ~ "No change"))

# summarise to get % of all children in each ethnicity/school year who changed BMI category
bmi_category_clinical_change_by_ethnicity_summary <- bmi_category_clinical_change_by_ethnicity |> 
  filter(category_change == T) |> 
  group_by(School_Year, Ethnicity_Hudda) |> 
  summarise(count = sum(count),
            denominator = mean(denominator)) |> 
  ungroup() |> 
  PHEindicatormethods::phe_proportion(x = count,
                                      n = denominator,
                                      multiplier = 100)

# calculate proportion of children within ethncicity/school year/bmi category who fit into each adjusted category
adjusted_bmi_by_ethnicity_bmi_category_clinical <- ncmp_summarise(ncmp_data_period_8,
                                                         .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical, BMI_Category_Clinical_Adjusted),
                                                         .by_denom = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical),
                                                         .include_oo = F,
                                                         .round_5 = T) |> 
  mutate(category_change = case_when(BMI_Category_Clinical == BMI_Category_Clinical_Adjusted ~ F,
                                     .default = T))
```

# Introduction

The National Child Measurement Program (NCMP) is crucial tool in monitoring prevalence of overweight and obesity in children in England. The height and weight of children in Reception (aged 4-5) and Year 6 (aged 10-11) are measured, and converted into a body mass index (BMI) score. Each child's weight status (underweight, healthy weight, overweight or obese) is then calculated with reference to the exclusively White British 1990 child growth reference population (UK90).

At a population level, BMI categories are assigned as such:

- **Underweight**: BMI less than the 2nd centile of the UK90 according to age and sex
- **Healthy weight**: BMI between the 2nd and 85th centiles
- **Overweight**: BMI between the 85th and 95th centiles
- **Obese**: BMI on or above the 95th centile

At a clinical level, BMI categories are assigned as such:

- **Underweight**: BMI less than the 2nd centile of the UK90 according to age and sex
- **Healthy weight**: BMI between the 2nd and 91st centiles
- **Overweight**: BMI between the 91st and 98th centiles
- **Obese**: BMI on or above the 98th centile

Previous research has indicated that BMI is not an accurate measure of body fatness in some ethnic groups[^1][^2], particularly South Asian and Black children. Ethnic-specific BMI adjustments for South Asian and Black children to improve estimation of body fat have been developed[^3] but are not currently applied in analysis of NCMP data.

[^1]: Claire M Nightingale, Alicja R Rudnicka, Chris G Owen, Derek G Cook, Peter H Whincup. (2010) Patterns of body size and adiposity among UK children of South Asian, black African–Caribbean and white European origin: Child Heart And health Study in England (CHASE Study). International Journal of Epidemiology, [https://doi.org/10.1093/ije/dyq180](https://doi.org/10.1093/ije/dyq180)

[^2]: Nightingale CM, Rudnicka AR, Owen CG, Donin AS, Newton SL, Furness CA, et al. (2013) Are Ethnic and Gender Specific Equations Needed to Derive Fat Free Mass from Bioelectrical Impedance in Children of South Asian, Black African-Caribbean and White European Origin? Results of the Assessment of Body Composition in Children Study. PLoS ONE [https://doi.org/10.1371/journal.pone.0076426](https://doi.org/10.1371/journal.pone.0076426)

[^3]: Hudda MT, Nightingale CM, Donin AS, et al. (2017) Body mass index adjustments to increase the validity of body fatness assessment in UK Black African and South Asian children. Int J Obes (Lond). [doi:10.1038/ijo.2017.75](doi:10.1038/ijo.2017.75)

# Demographics
```{r demographics}
all_records <- nrow(ncmp_data_period_8)

reception_count <- ncmp_data_period_8 |> 
  count(School_Year) |> 
  filter(School_Year == "Reception") |> 
  pull(n)

y6_count <- ncmp_data_period_8 |> 
  count(School_Year) |> 
  filter(School_Year == "Year 6") |> 
  pull(n)

school_year_ethnicity_gender_table <- ncmp_data_period_8 |> 
  count(Gender, School_Year, Ethnicity_Hudda) |> 
  ungroup() |> 
  group_by(Gender, School_Year) |> 
  mutate(denom = sum(n),
         pct = n/denom*100,
         display = paste0(n, " (", round(pct, 1), "%)")) |> 
  select(-denom, -n, -pct) |> 
  arrange(School_Year) |> 
  pivot_wider(names_from = c("Gender", "School_Year"),
              values_from = "display") |> 
  knitr::kable(col.names = c("Ethnicity", "Girls", "Boys", "Girls", "Boys")) |> 
  kableExtra::add_header_above(c(" " = 1,
                                 "Reception" = 2,
                                 "Year 6" = 2))

```

A total of `r all_records` records were included in the analysis - `r reception_count` Reception children (aged 4-5 years) and `r y6_count` Year 6 children (aged 10-11 years). Included children were those with valid measurements who were measured in the three year period `r periods[8]`.

Children were grouped according to the ethnicity recorded in their NCMP measurement, using the following ethnic groupings:

- White ('White British', 'White Irish' or 'any other White background')
- Black ('Black African' 'Black Caribbean' or 'any other Black background')
- South Asian ('Indian', 'Pakistani' or 'Bangladeshi')
- Mixed and Other ('Mixed Ethnicity', 'any other ethnic group', 'any other Asian background' or 'Chinese')
- Unknown ('Not stated' or 'NA')

```{r demographics_table}
school_year_ethnicity_gender_table
```

# Methods

Unadjusted BMI was used to calculate the overall and ethnic-specific prevalence of overweight and obesity. Prevalence was calculated using both population-monitoring and clinical thresholds. 

BMI was then adjusted for children from South Asian and Black backgrounds, using ethnic-specific BMI adjustments derived by Hudda et al.[^3]  

- For South Asian children, BMI is adjusted by a constant term of +1.12kg/m2 for boys and +1.07kg/m2 for girls.
- For Black children, BMI adjustment varies by age, sex and BMI, with the largest adjustments in younger children with higher BMI. These adjustments were applied using a regression model extrapolated from a table of measurements provided by Hudda et al.
 
After adjustment, BMI was recategorised using the LMS growth tool. Again, prevalence of overweight and obesity was calculated using both population-monitoring and clinical thresholds.

Unless otherwise stated, the presented results were calculated using population-monitoring thresholds.

# Results

## Overall prevalence

When BMI is not adjusted, a total of `r oo_r_unadj`% of Reception children and `r oo_y6_unadj`% of Year 6 children are classed as overweight or obese. Followig BMI adjustment in Black and South Asian children, these proportions rise to `r oo_r_adj`% of Reception children and `r oo_y6_adj`% of Year 6 children.

::: panel-tabset

## Reception

```{r overall_prevalence_r}
graph <- bmi_category_by_school_year|>
  filter(School_Year == "Reception",
         BMI_Category != "Overweight/Obese") |> 
  mutate(BMI_Category = factor(BMI_Category,
                               levels = c("Obese", "Overweight", "Healthy Weight", "Underweight")),
         bmi_adj = factor(bmi_adj,
                          levels = c("Adjusted BMI", "Unadjusted BMI")),
         tooltip = paste0(bmi_adj, "\n",
                          BMI_Category, "\n",
                          round(pct, 1), "% (", round(lower_ci, 1), " - ", round(upper_ci, 1), "%)", "\n")) |> 
  ggplot(aes(y = BMI_Category,
             x = pct,
             fill = bmi_adj)) +
  geom_col_interactive(aes(tooltip = tooltip),
                       position = "dodge") +
  geom_errorbar(aes(xmin = lower_ci,
                    xmax = upper_ci),
                position = position_dodge(0.9),
                width = 0.2) +
  scale_fill_manual(values = c(bcc_colours[2], bcc_colours[4])) +
  theme_minimal() +
  scale_x_continuous(expand = c(0,0),
                     label = scales::label_percent(scale = 1),
                     name = "Percentage of children") +
  scale_y_discrete(name = "BMI Category") +
  guides(fill = guide_legend(reverse = T,
                             title = ""),
         colour = "none")

girafe(ggobj = graph)
```

## Year 6

```{r overall_prevalence_y6}
graph <- bmi_category_by_school_year|>
  filter(School_Year == "Year 6",
         BMI_Category != "Overweight/Obese") |> 
  mutate(BMI_Category = factor(BMI_Category,
                               levels = c("Obese", "Overweight", "Healthy Weight", "Underweight")),
         bmi_adj = factor(bmi_adj,
                          levels = c("Adjusted BMI", "Unadjusted BMI")),
         tooltip = paste0(bmi_adj, "\n",
                          BMI_Category, "\n",
                          round(pct, 1), "% (", round(lower_ci, 1), " - ", round(upper_ci, 1), "%)", "\n")) |> 
  ggplot(aes(y = BMI_Category,
             x = pct,
             fill = bmi_adj)) +
  geom_col_interactive(aes(tooltip = tooltip),
                       position = "dodge") +
  geom_errorbar(aes(xmin = lower_ci,
                    xmax = upper_ci),
                position = position_dodge(0.9),
                width = 0.2) +
  scale_fill_manual(values = c(bcc_colours[2], bcc_colours[4])) +
  theme_minimal() +
  scale_x_continuous(expand = c(0,0),
                     label = scales::label_percent(scale = 1),
                     name = "Percentage of children") +
  scale_y_discrete(name = "BMI Category") +
  guides(fill = guide_legend(reverse = T,
                             title = ""),
         colour = "none")

girafe(ggobj = graph)
```
:::

## Ethnic-specific prevalence

Prior to adjustment, `r oo_r_unadj_b`% of Black Reception children and `r oo_y6_unadj_b`% of Black Year 6 children are classed as overweight or obese. After adjustment, these values decrease to `r oo_r_adj_b`% and `r oo_y6_adj_b`% respectively.

In South Asian children, before adjustment `r oo_r_unadj_sa`% of Reception children and `r oo_y6_unadj_sa`% of Year 6 children are classed as overweight or obese. Following adjustment, prevalence of overweight and obesity increases to `r oo_r_adj_sa`% and `r oo_y6_adj_sa`% respectively.

::: panel-tabset

## Reception

```{r ethnic_prevalence_r}
graph <- bmi_category_by_ethnicity |>
  filter(School_Year == "Reception",
         BMI_Category != "Overweight/Obese",
         Ethnicity_Hudda %in% c("Black", "South Asian")) |> 
  mutate(BMI_Category = factor(BMI_Category,
                               levels = c("Obese", "Overweight", "Healthy Weight", "Underweight")),
         bmi_adj = factor(bmi_adj,
                          levels = c("Adjusted BMI", "Unadjusted BMI")),
         tooltip = paste0(bmi_adj, "\n",
                          Ethnicity_Hudda, "\n", 
                          BMI_Category, "\n",
                          round(pct, 1), "% (", round(lower_ci, 1), " - ", round(upper_ci, 1), "%)", "\n")) |> 
  ggplot(aes(y = BMI_Category,
             x = pct,
             fill = bmi_adj)) +
  geom_col_interactive(aes(tooltip = tooltip),
                       position = "dodge") +
  geom_errorbar(aes(xmin = lower_ci,
                xmax = upper_ci),
                position = position_dodge(0.9),
                width = 0.2) +
  scale_fill_manual(values = c(bcc_colours[2], bcc_colours[4])) +
  facet_wrap(~Ethnicity_Hudda) +
  theme_minimal() +
  scale_x_continuous(expand = c(0,0),
                     label = scales::label_percent(scale = 1),
                     name = "Percentage of children") +
  scale_y_discrete(name = "BMI Category") +
  guides(fill = guide_legend(reverse = T,
                             title = ""),
         colour = "none")

girafe(ggobj = graph)
```

## Year 6

```{r ethnic_prevalence_y6}
graph <- bmi_category_by_ethnicity |>
  filter(School_Year == "Year 6",
         BMI_Category != "Overweight/Obese",
         Ethnicity_Hudda %in% c("Black", "South Asian")) |> 
  mutate(BMI_Category = factor(BMI_Category,
                               levels = c("Obese", "Overweight", "Healthy Weight", "Underweight")),
         bmi_adj = factor(bmi_adj,
                          levels = c("Adjusted BMI", "Unadjusted BMI")),
         tooltip = paste0(bmi_adj, "\n",
                          Ethnicity_Hudda, "\n", 
                          BMI_Category, "\n",
                          round(pct, 1), "% (", round(lower_ci, 1), " - ", round(upper_ci, 1), "%)", "\n")) |> 
  ggplot(aes(y = BMI_Category,
             x = pct,
             fill = bmi_adj)) +
  geom_col_interactive(aes(tooltip = tooltip),
                       position = "dodge") +
  geom_errorbar(aes(xmin = lower_ci,
                xmax = upper_ci),
                position = position_dodge(0.9),
                width = 0.2) +
  scale_fill_manual(values = c(bcc_colours[2], bcc_colours[4])) +
  facet_wrap(~Ethnicity_Hudda) +
  theme_minimal() +
  scale_x_continuous(expand = c(0,0),
                     label = scales::label_percent(scale = 1),
                     name = "Percentage of children") +
  scale_y_discrete(name = "BMI Category") +
  guides(fill = guide_legend(reverse = T,
                             title = ""),
         colour = "none")

girafe(ggobj = graph)
```
:::

## Weight status change

```{r weight_status_change_figures}

# Black 
wsc_b_r <- bmi_category_clinical_change_by_ethnicity_summary |> 
  filter(Ethnicity_Hudda == "Black",
         School_Year == "Reception") |>
  pull(value) |> 
  round(1)

wsc_b_y6 <- bmi_category_clinical_change_by_ethnicity_summary |> 
  filter(Ethnicity_Hudda == "Black",
         School_Year == "Year 6") |>
  pull(value) |> 
  round(1)

wsc_b_r_ov_hw <- adjusted_bmi_by_ethnicity_bmi_category_clinical |> 
  filter(Ethnicity_Hudda == "Black",
         School_Year == "Reception",
         BMI_Category_Clinical == "Overweight",
         BMI_Category_Clinical_Adjusted == "Healthy Weight") |>
  pull(pct) |> 
  round(1)

wsc_b_y6_ov_hw <- adjusted_bmi_by_ethnicity_bmi_category_clinical |> 
  filter(Ethnicity_Hudda == "Black",
         School_Year == "Year 6",
         BMI_Category_Clinical == "Overweight",
         BMI_Category_Clinical_Adjusted == "Healthy Weight") |>
  pull(pct) |> 
  round(1)

wsc_b_r_ob_ov <- adjusted_bmi_by_ethnicity_bmi_category_clinical |> 
  filter(Ethnicity_Hudda == "Black",
         School_Year == "Reception",
         BMI_Category_Clinical == "Obese",
         BMI_Category_Clinical_Adjusted == "Overweight") |>
  pull(pct) |> 
  round(1)

wsc_b_y6_ob_ov <- adjusted_bmi_by_ethnicity_bmi_category_clinical |> 
  filter(Ethnicity_Hudda == "Black",
         School_Year == "Year 6",
         BMI_Category_Clinical == "Obese",
         BMI_Category_Clinical_Adjusted == "Overweight") |>
  pull(pct) |> 
  round(1)

# South Asian

wsc_sa_r <- bmi_category_clinical_change_by_ethnicity_summary |> 
  filter(Ethnicity_Hudda == "South Asian",
         School_Year == "Reception") |>
  pull(value) |> 
  round(1)

wsc_sa_y6 <- bmi_category_clinical_change_by_ethnicity_summary |> 
  filter(Ethnicity_Hudda == "South Asian",
         School_Year == "Year 6") |>
  pull(value) |> 
  round(1)

wsc_sa_r_hw_ov <- adjusted_bmi_by_ethnicity_bmi_category_clinical |> 
  filter(Ethnicity_Hudda == "South Asian",
         School_Year == "Reception",
         BMI_Category_Clinical == "Healthy Weight",
         BMI_Category_Clinical_Adjusted == "Overweight") |>
  pull(pct) |> 
  round(1)

wsc_sa_y6_hw_ov <- adjusted_bmi_by_ethnicity_bmi_category_clinical |> 
  filter(Ethnicity_Hudda == "South Asian",
         School_Year == "Year 6",
         BMI_Category_Clinical == "Healthy Weight",
         BMI_Category_Clinical_Adjusted == "Overweight") |>
  pull(pct) |> 
  round(1)

wsc_sa_r_ov_ob <- adjusted_bmi_by_ethnicity_bmi_category_clinical |> 
  filter(Ethnicity_Hudda == "South Asian",
         School_Year == "Reception",
         BMI_Category_Clinical == "Overweight",
         BMI_Category_Clinical_Adjusted == "Obese") |>
  pull(pct) |> 
  round(1)

wsc_sa_y6_ov_ob <- adjusted_bmi_by_ethnicity_bmi_category_clinical |> 
  filter(Ethnicity_Hudda == "South Asian",
         School_Year == "Year 6",
         BMI_Category_Clinical == "Overweight",
         BMI_Category_Clinical_Adjusted == "Obese") |>
  pull(pct) |> 
  round(1)
```

Note: as the values in this section relate to individual children, clinical monitoring thresholds have been used to calculate weight status.

`r wsc_b_r`% of Reception children and `r wsc_b_y6`% of Year 6 children from Black backgrounds were classified as having a heavier weight status after ethnic-specific BMI adjustment. More specifically, `r wsc_b_r_ov_hw`% and `r wsc_b_y6_ov_hw`% of Reception and Year 6 children respectively were reclassified from 'Overweight' to 'Healthy Weight'. Similarly, `r wsc_b_r_ob_ov`% and `r wsc_b_y6_ob_ov`% of Reception and Year 6 children respectively were reclassified from 'Obese' to 'Overweight'.

`r wsc_sa_r`% of Reception children and `r wsc_sa_y6`% of Year 6 children from South Asian backgrounds were classified as having a heavier weight status after ethnic-specific BMI adjustment. More specifically, `r wsc_sa_r_hw_ov`% and `r wsc_sa_y6_hw_ov`% of Reception and Year 6 children respectively were reclassified from 'Healthy Weight' to 'Overweight'. Similarly, `r wsc_sa_r_ov_ob`% and `r wsc_sa_y6_ov_ob`% of Reception and Year 6 children respectively were reclassified from 'Overweight' to 'Obese'.

Further detail on movement between BMI categories before and after BMI adjustment is shown in the flowcharts below.

::: panel-tabset
## Black

::: panel-tabset

## Reception

```{r sankey_black_r}
bmi_adj_sankey_clinical(.data = adjusted_bmi_by_ethnicity_bmi_category_clinical,
               .Ethnicity_Hudda = "Black",
               .School_Year = "Reception")
```

## Year 6
```{r sankey_black_y6}
bmi_adj_sankey_clinical(.data = adjusted_bmi_by_ethnicity_bmi_category_clinical,
               .Ethnicity_Hudda = "Black",
               .School_Year = "Year 6")
```

:::

## South Asian

::: panel-tabset

## Reception
```{r sankey_south_asian_r}
bmi_adj_sankey_clinical(.data = adjusted_bmi_by_ethnicity_bmi_category_clinical,
               .Ethnicity_Hudda = "South Asian",
               .School_Year = "Reception")
```

## Year 6
```{r sankey_south_asian_y6}
bmi_adj_sankey_clinical(.data = adjusted_bmi_by_ethnicity_bmi_category_clinical,
               .Ethnicity_Hudda = "South Asian",
               .School_Year = "Year 6")
```
:::

:::