---
title: "NCMP Ethnic-specific BMI Adjustment Report"
date: last-modified
date-format: "MMM YYYY"
format: 
  html:
    grid:
      body-width: 900px
    embed-resources: true
toc: true
editor: source
execute: 
  echo: false
  warning: false
---

```{r setup}
library(tidyverse)
library(readxl)
library(plotly)
library(PHEindicatormethods)
library(bcctheme)

BMI_Category_Levels = c("Underweight", "Healthy Weight", "Overweight", "Obese")
bcc_colours <- unname(bcc_cols())
BMI_Category_colours <- c("Underweight" = bcc_colours[6],
                          "Healthy Weight" = bcc_colours[4],
                          "Overweight" = bcc_colours[5],
                          "Obese" = bcc_colours[1]) 

# function to round an integer to nearest 5
# for data suppression purposes: https://digital.nhs.uk/data-and-information/find-data-and-publications/statement-of-administrative-sources/methodological-changes/changes-to-the-national-child-measurement-programme-2019-20-publication-content-and-disclosure-control-methodology#simple-calculations
round_5 <- function(x) {
  return(round(x / 5) * 5)
}

# function to summarise a patient level dataset by the columns given
# .by_count contains columns to summarise by
# .by_denom contains columns to create denominator
# returns a dataframe containing count, denominator and percentage with 95% CIs
# optionally include overweight/obese category (for BMI_Category data only)
# optionally recalculate percentage using count and denom rounded to nearest 5
ncmp_summarise <- function(.data, 
                           .by_count, 
                           .by_denom, 
                           .include_oo,
                           .round_5){
  if(.include_oo == TRUE){
    temp <- .data |> 
      dplyr::summarise(count = n(), .by = {{ .by_count }}) |> 
      dplyr::mutate(denominator = sum(count), .by = {{ .by_denom }})
    
    temp_oo <- .data |> 
      dplyr::mutate(BMI_Category = dplyr::case_when(BMI_Category == "Obese" ~ "Overweight/Obese",
                                      BMI_Category == "Overweight" ~ "Overweight/Obese",
                                      .default = BMI_Category)) |> 
      dplyr::summarise(count = n(), .by = {{ .by_count }}) |> 
      dplyr::mutate(denominator = sum(count), .by = {{ .by_denom }}) |> 
      dplyr::filter(BMI_Category == "Overweight/Obese")
    
    temp <- rbind(temp, temp_oo)
    
    if(.round_5 == TRUE){
      temp <- PHEindicatormethods::phe_proportion(temp,
                                                  x = count,
                                                  n = denominator,
                                                  multiplier = 100) |> 
        dplyr::mutate(value = (round_5(count)/round_5(denominator))*100) |> 
        dplyr::rename(pct = value,
               lower_ci = lowercl,
               upper_ci = uppercl)
    }
    
    if(.round_5 == FALSE){
      temp <- PHEindicatormethods::phe_proportion(temp,
                                                  x = count,
                                                  n = denominator,
                                                  multiplier = 100) |> 
        dplyr::rename(pct = value,
               lower_ci = lowercl,
               upper_ci = uppercl)
    }
  }
  
  if(.include_oo == FALSE){
    temp <- .data |> 
      dplyr::summarise(count = n(), .by = {{ .by_count }}) |> 
      dplyr::mutate(denominator = sum(count), .by = {{ .by_denom }})
    
    if(.round_5 == TRUE){
      temp <- PHEindicatormethods::phe_proportion(temp,
                                                  x = count,
                                                  n = denominator,
                                                  multiplier = 100) |> 
        dplyr::mutate(value = (round_5(count)/round_5(denominator))*100) |> 
        dplyr::rename(pct = value,
               lower_ci = lowercl,
               upper_ci = uppercl)
    }
    
    if(.round_5 == FALSE){
      temp <- PHEindicatormethods::phe_proportion(temp,
                                                  x = count,
                                                  n = denominator,
                                                  multiplier = 100) |> 
        dplyr::rename(pct = value,
               lower_ci = lowercl,
               upper_ci = uppercl)
    }
  }
  temp
}

# function to generate an interactive sankey diagram showing movement between BMI classes
# .Ethnicity_Hudda is ethnicity to plot
#. School_Year is school year to plot
bmi_adj_sankey <- function(.data,
                           .Ethnicity_Hudda,
                           .School_Year){
  
  sankey_links <- .data |>
    dplyr::filter(Ethnicity_Hudda == .Ethnicity_Hudda,
           School_Year == .School_Year) |> 
    dplyr::select(BMI_Category, BMI_Category_Adjusted, count, pct, lower_ci, upper_ci) |>
    dplyr::rename(source = BMI_Category,
           target = BMI_Category_Adjusted,
           value  = count) |>
    dplyr::mutate(
      linkgroup = source,
      linklabel = dplyr::case_when(
        source == "Underweight"     & target == "Underweight"     ~ "Remained underweight",
        source == "Healthy Weight"  & target == "Healthy Weight"  ~ "Remained healthy weight",
        source == "Overweight"      & target == "Overweight"      ~ "Remained overweight",
        source == "Obese"           & target == "Obese"           ~ "Remained obese",
        .default = paste0("Moved to ", str_to_lower(target))),
      source = paste0(source, "_unadj"),
      target = paste0(target, "_adj"),
      colour = dplyr::case_when(
        linkgroup == "Underweight"    ~ paste0(BMI_Category_colours["Underweight"], "99"),
        linkgroup == "Healthy Weight" ~ paste0(BMI_Category_colours["Healthy Weight"], "99"),
        linkgroup == "Overweight"     ~ paste0(BMI_Category_colours["Overweight"], "99"),
        linkgroup == "Obese"          ~ paste0(BMI_Category_colours["Obese"], "99"))) |>
    dplyr::filter(value > 7)
  
  # Explicit node order (left then right)
  ordered_nodes <- c("Underweight_unadj", "Healthy Weight_unadj", "Overweight_unadj", "Obese_unadj",
                     "Underweight_adj", "Healthy Weight_adj", "Overweight_adj", "Obese_adj")
  
  # Build nodes with consistent labels and colours
  sankey_nodes <- tibble::tibble(name = ordered_nodes,
                         # category without suffix, for color mapping
                         nodegroup = sub("_(unadj|adj)$", "", ordered_nodes),
                         colour = dplyr::case_when(
                           nodegroup == "Underweight"    ~ BMI_Category_colours["Underweight"],
                           nodegroup == "Healthy Weight" ~ BMI_Category_colours["Healthy Weight"],
                           nodegroup == "Overweight"     ~ BMI_Category_colours["Overweight"],
                           nodegroup == "Obese"          ~ BMI_Category_colours["Obese"]))
  
  # Labels for nodes
  node_labels <- sub("_(unadj|adj)$", "", sankey_nodes$name)
  
  # Map links to node indices based on explicit node order
  sankey_links$IDsource <- match(sankey_links$source, sankey_nodes$name) - 1
  sankey_links$IDtarget <- match(sankey_links$target, sankey_nodes$name) - 1
  
  # Plot
  plotly::plot_ly(type = "sankey",
          arrangement = "snap",
          orientation = "h",
          node = list(label = node_labels,          # length 8; no recycling
                      color = sankey_nodes$colour,
                      pad = 15,
                      thickness = 20,
                      line = list(color = "black", width = 0.5),
                      hoverinfo = "none"),
          link = list(source = sankey_links$IDsource,
                      target = sankey_links$IDtarget,
                      value  = sankey_links$value,
                      label  = sankey_links$linklabel,
                      color  = sankey_links$colour,
                      customdata = sprintf("%.1f%% (%.1f%% - %.1f%%)", sankey_links$pct, sankey_links$lower_ci, sankey_links$upper_ci),
                      hovertemplate = paste0("%{label}<br>%{customdata}<extra></extra>"))) |> 
    add_annotations(x = 0, y = 1.01, xref = "paper", yref = "paper",
                    text = "Unadjusted BMI score", showarrow = FALSE,
                    xanchor = "left", yanchor = "bottom",
                    font = list(size = 16, color = "black")) |> 
    add_annotations(x = 1, y = 1.01, xref = "paper", yref = "paper",
                    text = "Adjusted BMI score", showarrow = FALSE,
                    xanchor = "right", yanchor = "bottom",
                    font = list(size = 16, color = "black")) |> 
    layout(margin = list(t = 40)) |> 
    config(modeBarButtonsToRemove = c('zoom', 'pan', 'select', 'zoomIn', 'zoomOut', 'lasso2d', 'autoScale', 'resetScale'))
}
```

```{r data_import}
ncmp_data <- read_excel("../data/ncmp_data_adjusted_bmi.xlsx") |> 
  rename(BMI_z_Adjusted = SDS_BMI) |> 
  mutate(BMI_p_Adjusted = pnorm(BMI_z_Adjusted),
         BMI_Category_Adjusted = case_when(BMI_p_Adjusted >= 0.95 ~ "Obese",
                                           BMI_p_Adjusted >= 0.85 ~ "Overweight",
                                           BMI_p_Adjusted > 0.02 ~ "Healthy Weight",
                                           BMI_p_Adjusted <= 0.02 ~ "Underweight"),
         BMI_Category_Clinical_Adjusted = case_when(BMI_z_Adjusted >= 6/3 ~ "Obese",
                                                    BMI_z_Adjusted >= 4/3 ~ "Overweight",
                                                    BMI_z_Adjusted > -6/3 ~ "Healthy Weight",
                                                    BMI_z_Adjusted <= -6/3 ~ "Underweight"))

# Group into three year combined periods ----------------------------------

# put years into a vector
years <- unique(ncmp_data$Year)

# remove 2020/2021
years <- years[!years %in% c("2020/2021")]

# create df with three years of data (first three years i.e. oldest)
ncmp_data_period_1 <- ncmp_data |> 
  filter(Year == years[1]|Year == years[2]|Year == years[3]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[1]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[3])))

# same again but with next three year period
ncmp_data_period_2 <- ncmp_data |> 
  filter(Year == years[2]|Year == years[3]|Year == years[4]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[2]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[4])))

# and so on
ncmp_data_period_3 <- ncmp_data |> 
  filter(Year == years[3]|Year == years[4]|Year == years[5]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[3]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[5])))

ncmp_data_period_4 <- ncmp_data |> 
  filter(Year == years[4]|Year == years[5]|Year == years[6]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[4]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[6])))

ncmp_data_period_5 <- ncmp_data |> 
  filter(Year == years[5]|Year == years[6]|Year == years[7]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[5]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[7])))

ncmp_data_period_6 <- ncmp_data |> 
  filter(Year == years[6]|Year == years[7]|Year == years[8]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[6]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[8])))

ncmp_data_period_7 <- ncmp_data |> 
  filter(Year == years[7]|Year == years[8]|Year == years[9]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[7]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[9])))

# most recent period
ncmp_data_period_8 <- ncmp_data |> 
  filter(Year == years[8]|Year == years[9]|Year == years[10]) |> 
  mutate(Year = paste0(sub("(20)(.*?)(20)", "\\1\\2", years[8]), " to ", sub("(20)(.*?)(20)", "\\1\\2", years[10])))

# bind all into one df
ncmp_data_3_years_combined <- bind_rows(ncmp_data_period_1,
                                        ncmp_data_period_2,
                                        ncmp_data_period_3,
                                        ncmp_data_period_4,
                                        ncmp_data_period_5,
                                        ncmp_data_period_6,
                                        ncmp_data_period_7,
                                        ncmp_data_period_8)

periods <- unique(ncmp_data_3_years_combined$Year)

# get a list of periods in format 20XX-XX
# where most recent period is periods[8]
periods_short <- sub("/([A-Za-z0-9]+( [A-Za-z0-9]+)+)/", "-", unique(ncmp_data_3_years_combined$Year))
```

```{r create_dfs}
# Find prevalences before and after adjustment (population) ----------------------------

# summarise unadjusted BMI categories by ethnicity

bmi_category_unadj_by_ethnicity <- ncmp_summarise(ncmp_data_period_8,
                                                  .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category),
                                                  .by_denom = c(School_Year, Ethnicity_Hudda),
                                                  .include_oo = T,
                                                  .round_5 = T) |> 
  mutate(bmi_adj = "Unadjusted")

# summarise adjusted BMI categories by ethnicity

bmi_category_adj_by_ethnicity <- ncmp_data_period_8 |> 
  select(-BMI_Category) |> 
  rename(BMI_Category = BMI_Category_Adjusted) |> 
  ncmp_summarise(.by_count = c(School_Year, Ethnicity_Hudda, BMI_Category),
                 .by_denom = c(School_Year, Ethnicity_Hudda),
                 .include_oo = T,
                 .round_5 = T) |> 
  mutate(bmi_adj = "Adjusted")

# combine into one table

bmi_category_by_ethnicity = rbind(bmi_category_adj_by_ethnicity,
                                  bmi_category_unadj_by_ethnicity)

# Find prevalences before and after adjustment (clinical) -----------------

# repeat for clinical categories

# summarise unadjusted BMI categories by ethnicity

bmi_category_clinical_unadj_by_ethnicity <- ncmp_summarise(ncmp_data_period_8,
                                                           .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical),
                                                           .by_denom = c(School_Year, Ethnicity_Hudda),
                                                           .include_oo = F,
                                                           .round_5 = T) |> 
  mutate(bmi_adj = "Unadjusted")

# summarise adjusted BMI categories by ethnicity

bmi_category_clinical_adj_by_ethnicity <- ncmp_data_period_8 |> 
  ncmp_summarise(.by_count = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical),
                 .by_denom = c(School_Year, Ethnicity_Hudda),
                 .include_oo = F,
                 .round_5 = T) |> 
  mutate(bmi_adj = "Adjusted")

# combine into one table

bmi_category_clinical_by_ethnicity = rbind(bmi_category_clinical_adj_by_ethnicity,
                                           bmi_category_clinical_unadj_by_ethnicity)

# Find % children changed category (population) ---------------------------

# calculate proportion of children within ethnicity/school year who fit into each BMI category/adjusted category combo
bmi_category_change_by_ethnicity <- ncmp_summarise(ncmp_data_period_8,
               .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category, BMI_Category_Adjusted),
               .by_denom = c(School_Year, Ethnicity_Hudda),
               .include_oo = F,
               .round_5 = T) |> 
  mutate(category_change = case_when(BMI_Category == BMI_Category_Adjusted ~ F,
                                     .default = T))

# add cols with number of "steps" and direction of change
bmi_category_change_by_ethnicity <- bmi_category_change_by_ethnicity |> 
  mutate(BMI_Category = factor(BMI_Category, levels = BMI_Category_Levels, ordered = T),
         BMI_Category_Adjusted = factor(BMI_Category_Adjusted, levels = BMI_Category_Levels, ordered = T),
         change_steps = as.numeric(unclass(BMI_Category_Adjusted) - unclass(BMI_Category)),
         change_direction = case_when(change_steps < 0 ~ "Lighter",
                                      change_steps > 0 ~ "Heavier",
                                      change_steps == 0 ~ "No change"))

# summarise to get % of all children in each ethnicity/school year who changed BMI category
bmi_category_change_by_ethnicity_summary <- bmi_category_change_by_ethnicity |> 
  filter(category_change == T) |> 
  group_by(School_Year, Ethnicity_Hudda) |> 
  summarise(count = sum(count),
            denominator = mean(denominator)) |> 
  ungroup() |> 
  PHEindicatormethods::phe_proportion(x = count,
                                      n = denominator,
                                      multiplier = 100)

# calculate proportion of children within ethncicity/school year/bmi category who fit into each adjusted category
adjusted_bmi_by_ethnicity_bmi_category <- ncmp_summarise(ncmp_data_period_8,
                                                         .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category, BMI_Category_Adjusted),
                                                         .by_denom = c(School_Year, Ethnicity_Hudda, BMI_Category),
                                                         .include_oo = F,
                                                         .round_5 = T) |> 
  mutate(category_change = case_when(BMI_Category == BMI_Category_Adjusted ~ F,
                                     .default = T))

# Find % children changed category (clinical) -----------------------------


# repeat for clinical category
# calculate proportion of children within ethnicity/school year who fit into each BMI category/adjusted category combo
bmi_category_clinical_change_by_ethnicity <- ncmp_summarise(ncmp_data_period_8,
                                                   .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical, BMI_Category_Clinical_Adjusted),
                                                   .by_denom = c(School_Year, Ethnicity_Hudda),
                                                   .include_oo = F,
                                                   .round_5 = T) |> 
  mutate(category_change = case_when(BMI_Category_Clinical == BMI_Category_Clinical_Adjusted ~ F,
                                     .default = T))

# add cols with number of "steps" and direction of change
bmi_category_clinical_change_by_ethnicity <- bmi_category_clinical_change_by_ethnicity |> 
  mutate(BMI_Category_Clinical = factor(BMI_Category_Clinical, levels = BMI_Category_Levels, ordered = T),
         BMI_Category_Clinical_Adjusted = factor(BMI_Category_Clinical_Adjusted, levels = BMI_Category_Levels, ordered = T),
         change_steps = as.numeric(unclass(BMI_Category_Clinical_Adjusted) - unclass(BMI_Category_Clinical)),
         change_direction = case_when(change_steps < 0 ~ "Lighter",
                                      change_steps > 0 ~ "Heavier",
                                      change_steps == 0 ~ "No change"))

# summarise to get % of all children in each ethnicity/school year who changed BMI category
bmi_category_clinical_change_by_ethnicity_summary <- bmi_category_clinical_change_by_ethnicity |> 
  filter(category_change == T) |> 
  group_by(School_Year, Ethnicity_Hudda) |> 
  summarise(count = sum(count),
            denominator = mean(denominator)) |> 
  ungroup() |> 
  PHEindicatormethods::phe_proportion(x = count,
                                      n = denominator,
                                      multiplier = 100)

# calculate proportion of children within ethncicity/school year/bmi category who fit into each adjusted category
adjusted_bmi_by_ethnicity_bmi_category_clinical <- ncmp_summarise(ncmp_data_period_8,
                                                         .by_count = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical, BMI_Category_Clinical_Adjusted),
                                                         .by_denom = c(School_Year, Ethnicity_Hudda, BMI_Category_Clinical),
                                                         .include_oo = F,
                                                         .round_5 = T) |> 
  mutate(category_change = case_when(BMI_Category_Clinical == BMI_Category_Clinical_Adjusted ~ F,
                                     .default = T))
```

# Introduction

The National Child Measurement Program (NCMP) is crucial tool in monitoring prevalence of overweight and obesity in children in England. The height and weight of children in Reception (aged 4-5) and Year 6 (aged 10-11) are measured, and converted into a body mass index (BMI) score. Each child's weight status (underweight, healthy weight, overweight or obese) is then calculated with reference to the exclusively White British 1990 child growth reference population (UK90).

At a population level, BMI categories are assigned as such:

- **Underweight**: BMI less than the 2nd centile of the UK90 according to age and sex
- **Healthy weight**: BMI between the 2nd and 85th centiles
- **Overweight**: BMI between the 85th and 95th centiles
- **Obese**: BMI on or above the 95th centile

At an individual child level, BMI categories are assigned as such:

- **Underweight**: BMI less than the 2nd centile of the UK90 according to age and sex
- **Healthy weight**: BMI between the 2nd and 91st centiles
- **Overweight**: BMI between the 91st and 98th centiles
- **Obese**: BMI on or above the 98th centile

Previous research has indicated that BMI is not an accurate measure of body fatness in some ethnic groups[^1][^2], particularly South Asian and Black children. Ethnic-specific BMI adjustments for South Asian and Black children to improve estimation of body fat have been developed[^3] but are not currently applied in analysis of NCMP data.

[^1]: Claire M Nightingale, Alicja R Rudnicka, Chris G Owen, Derek G Cook, Peter H Whincup. (2010) Patterns of body size and adiposity among UK children of South Asian, black African–Caribbean and white European origin: Child Heart And health Study in England (CHASE Study). International Journal of Epidemiology, [https://doi.org/10.1093/ije/dyq180](https://doi.org/10.1093/ije/dyq180)

[^2]: Nightingale CM, Rudnicka AR, Owen CG, Donin AS, Newton SL, Furness CA, et al. (2013) Are Ethnic and Gender Specific Equations Needed to Derive Fat Free Mass from Bioelectrical Impedance in Children of South Asian, Black African-Caribbean and White European Origin? Results of the Assessment of Body Composition in Children Study. PLoS ONE [https://doi.org/10.1371/journal.pone.0076426](https://doi.org/10.1371/journal.pone.0076426)

[^3]: Hudda MT, Nightingale CM, Donin AS, et al. (2017) Body mass index adjustments to increase the validity of body fatness assessment in UK Black African and South Asian children. Int J Obes (Lond). [doi:10.1038/ijo.2017.75](doi:10.1038/ijo.2017.75)

# Sankey

::: panel-tabset
## Black, Reception

```{r sankey_black_r}
bmi_adj_sankey(.data = adjusted_bmi_by_ethnicity_bmi_category,
               .Ethnicity_Hudda = "Black",
               .School_Year = "Reception")
```

## Black, Year 6
```{r sankey_black_y6}
bmi_adj_sankey(.data = adjusted_bmi_by_ethnicity_bmi_category,
               .Ethnicity_Hudda = "Black",
               .School_Year = "Year 6")
```

## South Asian, Reception
```{r sankey_south_asian_r}
bmi_adj_sankey(.data = adjusted_bmi_by_ethnicity_bmi_category,
               .Ethnicity_Hudda = "South Asian",
               .School_Year = "Reception")
```

## South Asian, Reception
```{r sankey_south_asian_y6}
bmi_adj_sankey(.data = adjusted_bmi_by_ethnicity_bmi_category,
               .Ethnicity_Hudda = "South Asian",
               .School_Year = "Year 6")
```
:::